Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-03-13 paper/paper.tex
last changes for sub2

% \title{Dark Control:\\
%        Towards a Unified Account of Default Mode Function\\
%        by Markov Decision Processes}
\title{The Default Mode Network is a Reinforcement Learning Agent!}

@@

\begin{abstract}
The default mode network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
Specifically, we answer the question whether the neurobiological properties of the DMN
collectively provide the computational building blocks necessary for a Markov Decision Process.
%
We argue that our formal account of DMN function naturally accommodates as special cases
previous interpretations based on
(1) predictive coding,
(2) semantic associations, and
(3) a sentinel role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.
We argue that DMN implication in many
of the most advanced human capacities
can be recast as prediction error minimization
informed by internally generated probabilistic simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
allowing maximization of action outcomes across multiple time-scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience instrumental to iteratively
optimize the behavior of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
the brain is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this previously proposed contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN regions
is linked to the reward circuitry (Fig. 2),
thus linking two literatures with currently scarce cross-references.
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments.
At this stage, we emphasize the importance of differentiating which levels of observation are at play in the present account. A process model is not solely intended to capture behavior of the agent, such as cognitive accounts of DMN function, but also the neurocomputational specifics of the agent. Henceforth, we will use ``inference`` when describing aspects of the statistical model, ``prediction`` when referring to the neurobiological implementation, and words like ``forecast`` or ``forsee`` when referring to the behavior of the agent.

@@

\section{Reinforcement learning control: a process model for DMN function}
We argue the outlined neurobiological properties
of the DMN regions
to be sufficient for implementing all components
of a full-fledged reinforcement-learning (RL) system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

@@

\paragraph*{Why Markov Decision Processes?} One may wonder whether MDP models
are applicable to something as complex as human behavior. The success of such
models in diverse domains are numerous. For instance, financial trading is largely a manifestation of strategic decision-making of interacting human agents. According to how the market responds, the agent incurs gain or loss as environmental feedback of the executed financial actions. Recent research on automatizing market exchanges by algorithmic trading has successfully used MDPs as a framework for modeling these elaborate behavioral dynamics  ~\citep{brazdil2017,yang15,yang14,yang12,dempster2006,hult2010, abergel2017}.
%
MDPs have also been effective as a behavioral model in robotics
\citep{ng2004,abbeel2004} and in challenging multistep strategy games
~\citep{mnih2015,silver2016mastering,pritzel2017neural}. An even recent work~\citep{robinowitz2018}
develops an MDP-based Theory of Mind for learning agents (the \textit{ToM-Net}).
The idea is to use-meta learning (learning-to-learn) to build strong priors
about the behavior of of a population of other agents.

@@

\paragraph*{Reinforcement learning in the brain.}
Model-free RL can be plausibly realized in the human brain \citep{doherty2015structure, daw2014algorithmic}.
Indeed,
it has been proposed \citep{gershman2015computational} that a core property of human intelligence is the improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
can be a direct mapping between model-free RL learning algorithms
and aspects of the brain.
The neurotransmitter dopamine could serve
as a 'teaching signal' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward-processing circuitry, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space, transition probabilities between states,
reward contingencies, etc.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
An agent might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a location
offering relevant resources.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}
% Model-based assumes knowledge of the environment in terms of the transition probabilities between states, the reward contingencies for state-action pairs, or any form of representation of the environment.
Such knowledge can be partly extracted from the environment: the agent infers a model of the world while learning to take optimal decisions based on the current representation of the environment. These methods learn what the effect is going to be of taking a particular action in a particular state. The result is an estimate of the underlying MDP which can then be either solved exactly or approximately, depending on the setting and what is feasible.

@@

The goal of the RL agent is then to successively update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted time horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning of longer-term agendas that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays
for higher reward outcomes.
Such an agent puts relatively more emphasis on rewards expected in
a more distant future.
Concretely,
rewards that are not expected to come within
% \footnote{\textbf{N.B.:} $\frac{1}{1-\gamma} = 1 + \gamma + \gamma^2 + \gamma^3 + \ldots$}
$\tau := 1/(1 - \gamma)$
time steps from the present point are ignored.
The complexity reduction by time discounting
alleviates the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the far future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics.
A brain-imaging experiment in humans for instance identified hippocampal signals that specifically preceded upcoming choice performance in prospective planning in new environments
\citep{kaplan2017neural}.
This would be a simple control algorithm by evaluating all legal actions and selecting the action with
highest expected cumulative rewards.
In MDPs, MC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
This is because expected consequences of action choices can be well evaluated although only a subset of the states are actually considered \citep{daw2014algorithmic}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}

@@

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.
The more familiar and predictable the current environment, the more brain resources may remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context. In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making has led the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing control of the organism in general. Among all parts of the DMN, the RTPJ is perhaps the most evident candidate for a network-switching relay that calibrates between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

@@

From the perspective of a model-free RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

@@

\item Experiment (Humans): We hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits. Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced-back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.

@@

In sum,
recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects from the prediction coding hypothesis.
The present MDP account of DMN function may therefore
serve as a concrete implementation of
predictive coding ideas.
MDPs have the advantage of exposing an explicit
mechanisms for controlling the horizon of future considerations and for
how the internal representation of the world is updated,
as well as why certain predictions may be more relevant to the agent than others.

@@

% \paragraph{Links between FEP and RL.}
We finally note that FEP and active inference can be reframed in terms of
our model-free RL framework.
This becomes possible by recasting the Q-value function (i.e., expected long-term reward) maximized by the DMN to
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to using free-energy as a Q-value approximator for the MDP such that

>>>>

% \title{Dark Control:\\
%        Towards a Unified Account of Default Mode Function\\
%        by Markov Decision Processes}
\title{Dark Control: The Default Mode Network as a Reinforcement Learning Agent}

@@

\begin{abstract}
The default mode network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
The main purpose of the DMN, we argue,
may be to perform optimization based on Markov Decision Processes
through vicarious trial and error.
%
Our formal account of DMN function naturally accommodates as special cases
previous interpretations based on
(1) predictive coding,
(2) semantic associations, and
(3) a sentinel role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.
We argue that DMN implication in many
of the most advanced human capacities
can be recast as prediction error minimization
informed by internally generated probabilistic simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
allowing maximization of action outcomes across multiple time-scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Monte Carlo sampling.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience instrumental to iteratively
optimize the behavior of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
the brain is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this previously proposed contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN regions
is linked to the reward circuitry (Fig. 2),
thus linking two literatures with currently scarce cross-references.
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments.
At this stage, we emphasize the importance of differentiating which levels of observation are at play in the present account. A process model is not solely intended to capture behavior of the agent, such as cognitive accounts of DMN function, but also the neurocomputational specifics of the agent. Henceforth, we will use ``inference`` when describing aspects of the statistical model, ``prediction`` when referring to the neurobiological implementation, and words like ``forecast`` or ``forsee`` when referring to the behavior of the agent.
%
Notably, our account does not claim that the brain or neural activity in the DMN are completely identical with reinforcement learning algorithms. Rather, we advocate feedback-based learning strategies as an attractive alternative perspective
to understand DMN function.

@@

\section{Reinforcement learning control: a process model for DMN function}
We argue the outlined neurobiological properties
of the DMN regions
to be sufficient for implementing all components
of a full-fledged reinforcement learning (RL) system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

@@

\paragraph*{Why Markov Decision Processes?} One may wonder whether MDP models
are applicable to something as complex as human behavior. This class of
reinforcement-learning models has had numerous successes in diverse domains. For instance, financial trading is largely a manifestation of strategic decision-making of interacting human agents. According to how the market responds, the agent incurs gain or loss as environmental feedback of the executed financial actions. Recent research on automatizing market exchanges by algorithmic trading has effectively deployed MDPs as a framework for modeling these elaborate behavioral dynamics  ~\citep{brazdil2017,yang15,yang14,yang12,dempster2006,hult2010, abergel2017}.
%
MDPs have also been effective as a behavioral model in robotics
\citep{ng2004,abbeel2004} and in challenging multistep strategy games
~\citep{mnih2015,silver2016mastering,pritzel2017neural}. An even recent work~\citep{robinowitz2018}
develops an MDP-based reasoning about future behavior
of other agents (the \textit{ToM-Net}).
The idea is to use meta-learning (learning to learn) to build strong priors
about the behavior of of a population of other agents.

@@

\paragraph*{Reinforcement learning in the brain.}
RL can be plausibly realized in the human brain \citep{doherty2015structure, daw2014algorithmic}.
Indeed,
it has been proposed \citep{gershman2015computational} that a core property of human intelligence is the improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
can be a mapping between algorithmic aspects
underlying model-free and model-based RL learning
and neurobiological aspects underlying brain function.
The neurotransmitter dopamine could serve
as a 'teaching signal' to better estimate value associations
and action policies by modulating
synaptic plasticity in the reward-processing circuitry, including the NAc.
In contrast, model-based RL would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space, transition probabilities between states,
reward contingencies, etc.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
An agent might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a location
offering relevant resources.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}
% Model-based assumes knowledge of the environment in terms of the transition probabilities between states, the reward contingencies for state-action pairs, or any form of representation of the environment.
Such knowledge can be partly extracted from the environment: the agent infers a model of the world while learning to take optimal decisions based on the current representation of the environment. These methods learn what the effect is going to be of taking a particular action in a particular state. The result is an estimate of the underlying MDP which can then be either solved exactly or approximately, depending on the setting and what is feasible.

@@

The goal of the RL agent is then to successively update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted time horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning of longer-term agendas that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays
for higher reward outcomes.
Such an agent puts relatively more emphasis on rewards expected in
a more distant future.
Concretely,
rewards that are not expected to occur only within
% \footnote{\textbf{N.B.:} $\frac{1}{1-\gamma} = 1 + \gamma + \gamma^2 + \gamma^3 + \ldots$}
a very large number of
time steps from the present point are ignored.
The complexity reduction by time discounting
alleviates the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the far future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics.
A brain-imaging experiment in humans for instance identified hippocampal signals that specifically preceded upcoming choice performance in prospective planning in new environments
\citep{kaplan2017neural}.
It would be a simple strategy to evaluate
all legal actions and selecting the action with
highest expected cumulative rewards.
In MDPs, MC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
This is because expected consequences of action choices can be well evaluated although only a subset of the states are actually considered \citep{daw2014algorithmic}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}

@@

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.
The more familiar and predictable the current environment, the more brain resources may remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context. In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making has led the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing intervention of the organism on the world in general. Among all parts of the DMN, the RTPJ is perhaps the most evident candidate for a network-switching relay that calibrates between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

@@

From the perspective of a RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

@@

\item Experiment (Humans): We hypothesize that the relevant time horizon is modulated by various factors such as age, acute stress and time-enduring impulsivity traits
\citep{luksys2009stress, haushofer2014psychology}.
Using a temporal discounting experiment, it can be quantified how the time horizon is affected at the behavioral level and then traced-back to its corresponding neural representation. Such experimental investigation can be designed to examine between-group and within-group effects (e.g., impulsive population like chronic gamblers or drug addicts); and brought in context with the participant’s age and personality traits.

@@

In sum,
recasting DMN function in terms of MDPs therefore naturally incorporates
a majority of aspects from the prediction coding hypothesis.
The present MDP account of DMN function may therefore
serve as a concrete implementation of
predictive coding ideas.
MDPs have the advantage of exposing an explicit
mechanisms for modulating the horizon of future considerations and for
how the internal representation of the world is updated,
as well as why certain predictions may be more relevant to the agent than others.

@@

% \paragraph{Links between FEP and RL.}
We finally note that FEP and active inference can be reframed in terms of
our RL framework.
This becomes possible by recasting the Q-value function (i.e., expected long-term reward) maximized by the DMN to
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to using free-energy as a Q-value approximator for the MDP such that
